---
title: "Inferring the direction of transmission"
author: "Oliver Ratmann"
date: "2018-10-21"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{Vignette Title}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r setup, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>"
)
``` 

## Introduction
[In the previous tutorial](Rakai.03.reconstruct_transmission_networks.html), we
inferred transmission networks from large scale *phyloscanner* output. 

**In this tutorial, we will identify phylogenetically highly-supported
source-recipient pairs within reconstructed transmission networks.** We will:

1. Classify phylogenetic linkages based on their phylogenetic support. 
2. Test phylogenetic inferences into the direction of transmission against
   available epidemiologic data.
   

## Setting up the analysis
We resume analysis from [the previous
tutorial](Rakai.03.reconstruct_transmission_networks.html), and start by loading
the pairs of individuals between whom linkage could not be excluded, and the
transmission networks that could be inferred amongst them:
```{r, include=TRUE, eval=FALSE, echo=TRUE, tidy=TRUE}
HOME			<- '/Users/Oliver/sandbox/DeepSeqProjects'
indir			<- file.path(HOME, 'RakaiPopSample_phyloscanner_analysis')
infile.pairs	<- file.path(indir,'phsc_analysis_of_dataset_S1_allpairs.rda')
infile.networks	<- file.path(indir,'phsc_analysis_of_dataset_S1_allnetworks.rda')
load( infile.pairs )	# loads rtp, rplkl, rpw
load( infile.networks )	# loads rtn, rtnn
```
Do make sure that the directory names above do not start with "~", because 
the names are not expanded in the scripts below. White space, or characters like
'-' are OK.

## Classify phylogenetic linkages
The data.table `rtnn` contains all information needed to classify linkages.
Following analysis of the Rakai population-based sample, we specify as cut-off
for strong support of phylogenetic linkage that the proportion of deep-sequence
phylogenies with close and adjacent subgraphs of two indivuals in the
most likely transmission chain exceeds 60% (`conf.cut`= 60%):
```{r, include=TRUE, eval=FALSE, echo=TRUE, tidy=TRUE}
conf.cut		<- 0.6				
rtnn[, PHYLOSCANNER_CLASSIFY:= NA_character_]
set(rtnn, rtnn[, which(is.na(PTY_RUN))], 'PHYLOSCANNER_CLASSIFY', 'insufficient deep sequence data for at least one individual')
set(rtnn, rtnn[, which(!is.na(PTY_RUN) & is.na(LINK_12) & is.na(LINK_21))], 'PHYLOSCANNER_CLASSIFY', 'ph unlinked pair')
set(rtnn, rtnn[, which(!is.na(POSTERIOR_SCORE_LINKED) & POSTERIOR_SCORE_LINKED<=conf.cut)], 'PHYLOSCANNER_CLASSIFY', 'unclear if pair ph linked or unlinked')
set(rtnn, rtnn[, which(!is.na(POSTERIOR_SCORE_LINKED) & POSTERIOR_SCORE_LINKED>conf.cut)], 'PHYLOSCANNER_CLASSIFY', 'ph linked pair direction not resolved')
set(rtnn, rtnn[, which(!is.na(POSTERIOR_SCORE_LINKED) & POSTERIOR_SCORE_LINKED>conf.cut & POSTERIOR_SCORE_12>conf.cut)], 'PHYLOSCANNER_CLASSIFY', 'ph linked pair direction 12')
set(rtnn, rtnn[, which(!is.na(POSTERIOR_SCORE_LINKED) & POSTERIOR_SCORE_LINKED>conf.cut & POSTERIOR_SCORE_21>conf.cut)], 'PHYLOSCANNER_CLASSIFY', 'ph linked pair direction 21')
```

We already noted [in the previous
tutorial](Rakai.03.reconstruct_transmission_networks.html) that highly-supported
phylogenetic linkage does not exclude the possibility that unsampled
intermediates may be present, even with deep sequence data. To better interpret
inferred, phylogenetically highly supported linkages, we add information by
gender:
```{r, include=TRUE, eval=FALSE, echo=TRUE, tidy=TRUE}
tmp		<- subset(rtp, select=c(ID1, ID2, ID1_SEX, ID2_SEX))
rtnn	<- merge(rtnn, tmp, by=c('ID1','ID2'))
tmp		<- subset(rtnn, ID1_SEX=='F' & ID2_SEX=='M')
setnames(tmp, colnames(tmp), gsub('xx','ID2',gsub('ID2','ID1',gsub('ID1','xx',gsub('xx','12',gsub('12','21',gsub('21','xx',colnames(tmp))))))))
set(tmp, NULL, 'PHYLOSCANNER_CLASSIFY', tmp[, gsub('xx','12',gsub('12','21',gsub('21','xx',PHYLOSCANNER_CLASSIFY)))])
rtnn	<- rbind(subset(rtnn, !(ID1_SEX=='F' & ID2_SEX=='M')), tmp)
rtnn[, PAIR_SEX:= paste0(ID1_SEX,ID2_SEX)]
rtp		<- subset(rtnn, !grepl('unlinked|insufficient',PHYLOSCANNER_CLASSIFY))
rtp[, table(PAIR_SEX)]
```
Thus, a considerable proportion of highly supported phylogenetic links were between two women, 
even though HIV-1 is extremely rarely transmitted sexually between women, 
suggesting that unsampled intermediates are likely present. From these numbers, we expect that the 
proportion of highly supported male-female links between whom there are likely unsampled intermediates could be 
up to 35.4%, because there are almost twice as many possible male-female combinations than female-female 
combinations.

## Validate phylogenetic inferences into the direction of transmission
These observations prompted us to focus on using deep-sequence data for
inferences into the direction of transmission at the population-level, and
within partially sampled networks. Our main idea is that source cases, that were
phylogenetically identified, likely are transmitters regardless of the presence
of unsampled intermediates. Therefore, as a whole they provide evidence into the
sources of epidemic spread. To validate our inferences, we considered
phylogenetically linked pairs of individuals with epidemiological and clinical
data on the direction of transmission. We start by loading these data:
```{r, include=TRUE, eval=FALSE, echo=TRUE, tidy=TRUE}
infile		<- '~/sandbox/DeepSeqProjects/RakaiPopSample_data/Dataset_S3.csv'
red			<- as.data.table(read.csv(infile))
setnames(red, c('MALE_ID','FEMALE_ID'), c('ID1','ID2'))
```

Let us compare the epidemiologic data against our phylogenetic inferences:
```{r, include=TRUE, eval=FALSE, echo=TRUE, tidy=TRUE}
#	select male-female pairs with highly supported direction
rtpd	<- subset(rtnn, ID1_SEX=='M' & ID2_SEX=='F' & grepl('direction 12|direction 21',PHYLOSCANNER_CLASSIFY))
set(rtpd, NULL, 'PHYLOSCANNER_CLASSIFY', rtpd[, gsub('ph linked pair direction 21','fm',gsub('ph linked pair direction 12','mf',PHYLOSCANNER_CLASSIFY))])
setnames(rtpd, 'PHYLOSCANNER_CLASSIFY', 'PHYSCANNER_DIR')
rtpd	<- subset(rtpd, select=c(ID1, ID2, PHYSCANNER_DIR))
rtpd		<- merge(rtpd, red, by=c('ID1','ID2'))
rtpd[, PHYSCANNER_DIR_CONSISTENT:= as.integer(PHYSCANNER_DIR==EPID_EVIDENCE_DIR)]
rtpd[, table(PHYSCANNER_DIR_CONSISTENT)]
```
Thus, the large majority of phylogenetic inferences in our validation panel are consistent with 
epidemiologic and clinical data.
  