---
title: "Dating trees with ancestral states (phyloscanner + treedater)"
author: "Oliver Ratmann"
date: "2018-10-21"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{Vignette Title}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r setup, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>"
)
options(width=400)
``` 

```{css, echo=FALSE}
pre code, pre, code {
  white-space: pre !important;
  overflow-x: scroll !important;
  word-break: keep-all !important;
  word-wrap: initial !important;
}
```

## Introduction
This vignette illustrates new code that I added to the `ordev` branch of
`phyloscanner` in order to date trees in *phyloscanner* format. 

The rationale and steps in the analysis below are as follows:

* Many LANL sequences do not have a sampling date available, but we still like
  to use them in parsimony-based ancestral state reconstructions
* We perform parsimony-based ancestral state reconstructions with all LANL
  sequences using *phyloscanner*
* Then we remove taxa without sampling dates. We cannot simply use
  `ape:::drop.tip` because the function returns a new tree structure, and so
  our previous ancestral state reconstructions are meaningless. Luckily,
  `phytools` has functionality to drop tips and extract clades of trees with
  ancestral state reconstructions, though the trees need to be in `SIMMAP`
  format. So all we need to do is to cast the *phyloscanner* trees into
  `SIMMAP` format, and then we can drop taxa. The end result is a
  tree without sampling dates, and with ancestral state reconstructions
  preserved.
* The next step is to date the tree, with `treedater`.
* Since `treedater` keeps the tree topology intact, the edge index in the
  dated tree is the same as the edge index in the undated tree. This means we
  can copy the branch lengths of the dated tree into the previous undated tree.
  The end result is a dated tree, with dropped tips without sampling
  information, and with ancestral state reconstructions.
* Now we can extract King County subgraphs, that are dated. These
  subgraphs correspond to transmission chains that circulate in King County.  
* Note that we have sampling dates for all sequences from King County, so the
  above steps retain the entire King County data.
* The next step will be to perform phylodynamic analysis on the data that
  corresponds to transmission chains in Seattle.


## Getting started
You need to set the following input arguments:
 ```{r, include=TRUE, eval=TRUE, echo=TRUE, tidy=TRUE}
#	
#	file inputs
#infile.seqinfo <- '~/Box Sync/OR_Work/Seattle/PHSKC-2019_07_24/sequences_meta.rds'
#NOTE: new file above not compatible with old data: cannot find sequence ids.

infile.seqinfo <- '~/Box Sync/OR_Work/Seattle/PHSKC-2018-07-09/sequences_meta.rds'	
indir.trees	<- '~/Box Sync/OR_Work/Seattle/analysis_190723/trees'	
indir.phsc	<- '~/Box Sync/OR_Work/Seattle/analysis_190723/phyloscanner_HSX'
```	

Make sure you have the ordev branch of phyloscannerR installed:
```{r, include=TRUE, eval=TRUE, echo=TRUE, tidy=FALSE}
#	required packages
require(data.table) 	
require(tidyverse) 	
require(ape) 	
require(phytools)
require(treedater)
require(phyloscannerR)
	
#	file inputs
infiles.trees <- data.table(F=list.files(indir.trees, pattern='rerooted\\.newick$', recursive=TRUE, full.names=TRUE))
infiles.phsc <- data.table(F=list.files(indir.phsc, pattern='workspace.rda$', full.names=TRUE, recursive=TRUE))	

#	input args needed for tree.dater
alignment.length <- 1100

#
# read Seattle sampling data 
# define initial guess on actual sampling date as midpoint 
# define upper/lower limits on actual sampling date
dseq <- readRDS(infile.seqinfo) 
dseq <- dseq %>% 
			select(seqID, newnum, type, seqy, seqm) %>%
			mutate( SEQ_DATE:= seqy + (seqm-1)/12 + 15/365,
					SEQ_DATE_LOWER:= seqy + (seqm-1)/12,
					SEQ_DATE_UPPER:= seqy + (seqm-1)/12 + 30/365
					) %>%
			select(-seqy, -seqm) %>%
			rename(TAXA= seqID, PID= newnum, TYPE=type)
dseq <- as.data.table(dseq)
```	

##	Code to date a phyloscanner tree
The following code chunk will:

* load a tree with ancestral states in phyloscanner format
* remove taxa without sampling dates
* date the tree

```{r, include=TRUE, eval=TRUE, echo=TRUE, tidy=FALSE}
#
#	read tree with ancestral state reconstructions in phyloscanner format
#
i<- 1
infile <- infiles.phsc[i,F]	
load(infile)
ph <- phyloscanner.trees[[1]][['tree']]
stopifnot( !any( ph$tip.label=='' ) )

#
#	drop tips without sequence dates, 
#	while conserving the ancestral state reconstructions
#

#	extract taxa names
dph <- data.table(	TAXA_LABEL=ph$tip.label,
		TAXA= gsub(	'^(PR/RT-[0-9]+).*','\\1',
				gsub('^[^_]+___(.*)','\\1',
				  ph$tip.label)),
		TAXA_ID= seq_along(ph$tip.label))
#	add Seattle sequence dates
dph <- merge(dph, dseq, by='TAXA', all.x=TRUE)
#	extract GenBank sequence dates from taxa names where possible 
dph[, GENBANK_ID:= gsub('^([^\\.]+)\\.([^\\.]+)\\.([^\\.]+)\\.([^\\.]+)\\.([^\\.]+)\\.([^\\.]+)$','\\5',TAXA)]
dph[, GENBANK_SEQDATE:= gsub('^([^\\.]+)\\.([^\\.]+)\\.([^\\.]+)\\.([^\\.]+)\\.([^\\.]+)\\.([^\\.]+)$','\\3',TAXA)]
set(dph, dph[, which(GENBANK_SEQDATE=='-' | !is.na(SEQ_DATE))],'GENBANK_SEQDATE',NA_character_)	
set(dph, NULL, 'GENBANK_SEQDATE_LOWER', dph[, as.numeric(GENBANK_SEQDATE) ])
set(dph, NULL, 'GENBANK_SEQDATE_UPPER', dph[, as.numeric(GENBANK_SEQDATE) + 364/365 ])	
set(dph, NULL, 'GENBANK_SEQDATE', dph[, as.numeric(GENBANK_SEQDATE) + 1/2 ])
tmp <- dph[, which(is.na(SEQ_DATE))]
set(dph, tmp, 'SEQ_DATE', dph[tmp, GENBANK_SEQDATE])
set(dph, tmp, 'SEQ_DATE_LOWER', dph[tmp, GENBANK_SEQDATE_LOWER])
set(dph, tmp, 'SEQ_DATE_UPPER', dph[tmp, GENBANK_SEQDATE_UPPER])
set(dph, NULL, c('GENBANK_SEQDATE','GENBANK_SEQDATE_LOWER','GENBANK_SEQDATE_UPPER'), NULL)	
#	drop tips 
dph.old <- subset(dph, select=c(TAXA, SEQ_DATE, SEQ_DATE_LOWER, SEQ_DATE_UPPER))	
tmp <- subset(dph, is.na(SEQ_DATE))[, TAXA_ID]
cat('Dropping tips without sampling date from ', infile,' n=', length(tmp), 'of Ntips=', Ntip(ph), '\n')
ph <- phyloscanner.to.simmap(ph)	
ph <- phytools:::drop.tip.simmap(ph, ph$tip.label[tmp])
ph <- simmap.to.phyloscanner(ph)


#
#	date tree
#	

#	make data.table of sequence sampling times
dph <- data.table(	TAXA_LABEL=ph$tip.label,
		TAXA= gsub('^(PR/RT-[0-9]+).*','\\1',
				gsub('^[^_]+___(.*)','\\1',
				  ph$tip.label)),
		TAXA_ID= seq_along(ph$tip.label))
dph <- merge(dph, dph.old, by= 'TAXA')
stopifnot( !any(is.na(dph$SEQ_DATE)) )
dph <- dph[order(TAXA_ID),]
#	get into format needed for tree.dater	
sampling.times.init <- dph$SEQ_DATE
names(sampling.times.init) <- dph$TAXA_LABEL
sampling.times.bounds <- as.data.frame(subset(dph, select=c(SEQ_DATE_LOWER, SEQ_DATE_UPPER)))
rownames(sampling.times.bounds) <- dph$TAXA_LABEL		
colnames(sampling.times.bounds) <- c('lower','upper')
#	date tree
ph.dated <- dater(	ph, 
					sampling.times.init, 
					alignment.length, 
					numStartConditions=1, 
					estimateSampleTimes=sampling.times.bounds )
stopifnot( all(  ph.dated$tip.label == ph$tip.label ) )
stopifnot( all( ph.dated$edge == ph$edge ) )
#	since the tree topology is unchanged, we can copy
#	the branch lenghts in units of time onto the original tree
#	that has the ancestral state reconstructions
ph$edge.length <- ph.dated$edge.length
```	

Let us also print and save the output:
```{r, include=TRUE, eval=FALSE, echo=TRUE, tidy=TRUE}
#
#	plot dated tree to spot obvious errors
#
outfile <- gsub('workspace\\.rda','annotated_dated_tree.pdf',infile)
tmp <- vector('list')
tmp[['tree']] <- ph
tmp[['tree']][['node.states']] <- tmp[['tree']][['mapped.edge']] <- tmp[['tree']][['maps']] <- NULL
attr(tmp[['tree']],'map.order') <- NULL
attr(tmp[['tree']],'class') <- 'phylo'
tmp[['read.counts']] <- rep(1, Ntip(ph))	
write.annotated.tree(tmp, outfile, format="pdf", pdf.scale.bar.width = 0.01, pdf.w = 40, pdf.hm = 0.2, verbose = FALSE)

#
#	save phyloscanner.tree
#
outfile <- gsub('workspace\\.rda','annotated_dated_tree.rda',infile)
save(ph, file=outfile)
```

##	Code to extract King County subgraphs
The following code chunk will:

* take a dated tree with ancestral state reconstructions
* extract all King County subgraphs

```{r, include=TRUE, eval=TRUE, echo=TRUE, tidy=FALSE}
# phyloscanner label for transmission chains among heterosexual individuals in King County 
host <- 'KCHSX'

# find the MRCAS of all King County subgraphs 
mrcas <- which( attr(ph, 'SUBGRAPH_MRCA') )
mrcas <- mrcas[ attr(ph, 'INDIVIDUAL')[mrcas]==host ]	
# convert tree to class simmap
ph <- phyloscanner.to.simmap(ph)
# extract subgraphs
subgraphs <- lapply(mrcas, function(mrca) extract.subgraph(ph, mrca))
subgraphs
```	

Let us also save the output:
```{r, include=TRUE, eval=FALSE, echo=TRUE, tidy=TRUE}
outfile <- gsub('_workspace',paste0('_subgraphs_',host),infile)
save(subgraphs, file=outfile)
```

##	Notes
The vignette was generated with: 
```{r, include=TRUE, eval=FALSE, echo=TRUE, tidy=FALSE}
require(rmarkdown) 	
setwd('~/git/Phyloscanner.R.utilities/vignettes')
infile <- 'Seattle.treedater.Rmd'
rmarkdown::render(infile)	
```		